---
title: "Genk overtime: verification + scheduled-vs-unplanned decomposition"
author: "Haroon Tharwat"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE}
library(readxl)
library(dplyr)
library(tidyr)

# >>> SET THIS PATH to where genk_cleaned.xlsx lives on your PC <<<
genk <- read_excel("genk_cleaned.xlsx")
```

## 1. Baseline sanity check (must match the paper exactly)

```{r baseline}
nrow(genk)                                   # expect 79352

genk %>% filter(afterhours_flag == 1) %>%
  summarise(n = n(),
            rate_pct = 100 * n() / nrow(genk),
            mean_ot  = mean(overtime_minutes),
            median_ot = median(overtime_minutes),
            p95_ot   = quantile(overtime_minutes, .95))
# expect: n 7729 | 9.7% | 60.3 | 39 | 197
```

## 2. The new variable: was the case PLANNED to end past its shift boundary?

Same construction data_cleaning.R anticipated (PlannedEndDT vs planned_shift_end).

```{r planned-over}
genk <- genk %>%
  mutate(planned_over = PlannedEndDT > planned_shift_end,
         pov_min = as.numeric(difftime(PlannedEndDT, planned_shift_end, units = "mins")),
         urgent  = grepl("niet|non", tolower(UrgencyType)))

genk %>% summarise(n = sum(planned_over, na.rm = TRUE),
                   pct = 100 * mean(planned_over, na.rm = TRUE))
# expect: 5812 | 7.3%

genk %>% group_by(urgent) %>%
  summarise(pct_planned_over = 100 * mean(planned_over, na.rm = TRUE))
# expect: elective (FALSE) 6.4% | urgent (TRUE) 12.8%
```

## 3. Severity: how far past the boundary were they planned?

```{r bins}
genk %>% filter(planned_over) %>%
  mutate(bin = cut(pov_min, c(0, 15, 30, 60, 120, Inf),
                   labels = c("1-15", "16-30", "31-60", "61-120", ">120"))) %>%
  count(bin, urgent) %>%
  pivot_wider(names_from = urgent, values_from = n,
              names_glue = "{ifelse(urgent,'urgent','elective')}")
# expect totals per bin: 1299 / 1162 / 1362 / 1113 / 876

genk %>% filter(planned_over) %>%
  summarise(median_min_over = median(pov_min), p95 = quantile(pov_min, .95))
# expect: 40 | 246
```

## 4. The funnel: planned crossings convert to real overtime

```{r funnel}
genk %>% filter(planned_over) %>%
  summarise(crossed = sum(afterhours_flag == 1, na.rm = TRUE),
            pct = 100 * mean(afterhours_flag == 1, na.rm = TRUE))
# expect: 3532 | 60.8%

genk %>% filter(!planned_over) %>%
  summarise(pct_overtime = 100 * mean(afterhours_flag == 1, na.rm = TRUE))
# expect: 5.7%  -> overtime risk is ~10x higher when the plan already crosses
```

## 5. Decomposition of the 7,729 overtime cases

```{r decomposition}
ot <- genk %>% filter(afterhours_flag == 1)

ot %>% group_by(planned_over) %>%
  summarise(n = n(),
            case_share = 100 * n() / nrow(ot),
            mean_ot = mean(overtime_minutes),
            median_ot = median(overtime_minutes),
            total_min = sum(overtime_minutes)) %>%
  mutate(minutes_share = 100 * total_min / sum(total_min))
# expect: planned TRUE  3532 | 45.7% | 82 | 58 | 62.1% of minutes
#         planned FALSE 4197 | 54.3% | 42 | 28 | 37.9% of minutes
```

## 6. Per room: which kind of overtime does each room produce?

```{r per-room}
genk %>% group_by(ActualOR) %>%
  summarise(n = n(),
            ot_rate = 100 * mean(afterhours_flag == 1, na.rm = TRUE),
            scheduled_rate = 100 * mean(planned_over & afterhours_flag == 1, na.rm = TRUE),
            unplanned_rate = 100 * mean(!planned_over & afterhours_flag == 1, na.rm = TRUE)) %>%
  arrange(desc(ot_rate)) %>% print(n = 18)

ot %>% filter(ActualOR == "GO10") %>%
  summarise(scheduled_share = 100 * mean(planned_over))
# expect: 81.5%  -> GO10's overtime is mostly written into the schedule
```

## 7. Concentration of overtime minutes (candidate addition)

```{r concentration}
tot_min <- sum(ot$overtime_minutes)
ot %>% group_by(ActualOR) %>%
  summarise(minutes = sum(overtime_minutes)) %>%
  mutate(share_pct = 100 * minutes / tot_min) %>%
  arrange(desc(share_pct)) %>% head(5)
# expect: GO10 19.0% (with 2.2% of caseload); GO10+GO11+GO05 together 36.4%
```

## 8. The two correlations (decision: which idle denominator to report)

```{r correlations}
room <- genk %>% group_by(ActualOR) %>%
  summarise(late = mean(start_diff > 0, na.rm = TRUE),
            otr  = mean(afterhours_flag == 1, na.rm = TRUE),
            gap  = mean(gap_time, na.rm = TRUE))

cor.test(room$late, room$otr, method = "spearman")   # expect rho -0.292, p 0.239 (unchanged)

cor.test(room$gap, room$otr, method = "spearman")    # expect rho 0.913  (all-cases OT rate)

sub <- genk %>% filter(!is.na(gap_time)) %>% group_by(ActualOR) %>%
  summarise(gap = mean(gap_time), otr = mean(afterhours_flag == 1))
cor.test(sub$gap, sub$otr, method = "spearman")      # expect rho 0.8906 (your original run)

no10 <- room %>% filter(ActualOR != "GO10")
cor.test(no10$gap, no10$otr, method = "spearman")    # expect rho ~0.90
# -> the draft's "excluding OR10, the pattern did not hold" is refuted either way
```

## 9. OPTIONAL - only if we decide to run it (Niels's overlap question)

```{r overlap-optional, eval=FALSE}
is_overlap <- function(s1, e1, s2, e2) (s1 < e2) & (e1 > s2)

ne <- genk %>% filter(urgent, !is.na(ActualOR), !is.na(ORIn), !is.na(OROut)) %>%
  select(NE_OR = ActualOR, NE_start = ORIn, NE_end = OROut)
e  <- genk %>% filter(!urgent, !is.na(PlannedOR), !is.na(PlannedStartDT), !is.na(PlannedEndDT)) %>%
  select(E_ID = PatientID, E_OR = PlannedOR, E_start = PlannedStartDT, E_end = PlannedEndDT)

overlapped_ids <- ne %>% inner_join(e, by = c("NE_OR" = "E_OR"),
                                    relationship = "many-to-many") %>%
  filter(is_overlap(NE_start, NE_end, E_start, E_end)) %>% distinct(E_ID)

genk %>% filter(!urgent) %>%
  mutate(overlapped = PatientID %in% overlapped_ids$E_ID) %>%
  group_by(overlapped) %>%
  summarise(n = n(), overtime_rate = 100 * mean(afterhours_flag == 1, na.rm = TRUE))
# would answer: do overlap-affected electives run more overtime?
```
