---
title: "Planned versus Unplanned Overtime - Genk"
subtitle: Leveraging Data Analytics for Efficient OR Planning at ZOL
date: "August 2026"
output:
  pdf_document:
    latex_engine: xelatex
    toc: false
    keep_tex: true
  html_document:
    toc: false
    df_print: paged
geometry: margin=1in
header-includes:
  - |
    \usepackage{graphicx}
    \usepackage{booktabs}
    \usepackage{float}
    \usepackage{titling}

    % Logo's bovenaan
    \pretitle{%
      \begin{center}
      \includegraphics[width=4cm]{figures/BINF.png}\hfill
      \includegraphics[width=6cm]{figures/PAIH.png}\\[2cm]
      \huge
    }
    \posttitle{\end{center}}
---

\begin{center}
\vspace{2cm}

Niels MARTIN\\
niels.martin@uhasselt.be\\[0.5em]
Maxim RIEBUS\\
maxim.riebus@uhasselt.be\\[0.5em]
Haroon THARWAT\\
haroon.tharwat@uhasselt.be\\

\vspace{6cm}

\textbf{Hasselt University} \\
Faculty of Business Economics \\
Research Group Business Informatics \\
Process Analytics in Healthcare
\end{center}

\newpage
\tableofcontents
\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

knitr::opts_chunk$set(out.width = "70%")
knitr::opts_chunk$set(fig.align = "center")
knitr::opts_chunk$set(fig.asp = 0.618)

kable <- function(x, ...) {
  knitr::kable(x, booktabs = T, format = "latex",  position = "H", ...)
}

knitr::opts_chunk$set(fig.pos = "h", out.extra = "")

library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)
library(pander)
library(scales)
```

```{r data import, include=FALSE}
genk <- read_excel("genk_cleaned.xlsx")
```

```{r derive, include=FALSE}
genk <- genk %>%
  mutate(
    planned_over = PlannedEndDT > planned_shift_end,
    pov_min      = as.numeric(difftime(PlannedEndDT, planned_shift_end, units = "mins")),
    urgency_lbl  = ifelse(grepl("niet|non", tolower(UrgencyType)), "Urgent", "Elective")
  )
ot <- genk %>% filter(afterhours_flag == 1)
```

# Introduction

This document extends the in-depth analysis for campus Genk with one further question: to what extent does the surgical schedule itself already plan cases beyond the end of their staffed shift, and how does this planned component relate to the overtime we observe? The analysis uses the same cleaned dataset as the previous reports, and overtime keeps its established definition: a case whose room-out time falls after the end of its assigned shift. Distinguishing overtime that was written into the schedule from overtime that arose during execution matters operationally, because the two call for different responses.

# Starting point

As a starting point, Table \ref{tab:baseline} reproduces the headline overtime figures. They are identical to those reported in the in-depth analysis: 7,729 of 79,352 cases (9.7%) ran past the end of their assigned shift, with a mean overtime of 60.3 minutes, a median of 39 minutes, and a 95th percentile of 197 minutes.

```{r baseline, table.pos='H'}
ot %>%
  summarise(
    `Surgeries with overtime` = n(),
    `Percentage` = round(100 * n() / nrow(genk), 1),
    `Mean Overtime` = round(mean(overtime_minutes), 1),
    `Median Overtime` = round(median(overtime_minutes), 1),
    `P95 Overtime` = round(quantile(overtime_minutes, .95), 1)
  ) %>%
  pander(split.table = 120,
         caption = "Overall share and duration of overtime surgeries (Genk). \\label{tab:baseline}")
```

# How often does the schedule plan beyond the shift boundary?

A case is classified as a planned crossing when its planned end time falls after the end of its planned shift. This applies to 5,812 cases, or 7.3% of all surgeries: 6.4% of elective cases and 12.8% of urgent cases. Planning beyond the boundary is therefore not an occasional exception but a structural feature of the schedule, and it is roughly twice as common for urgent as for elective work. Table \ref{tab:pov} summarises the frequencies.

```{r pov, table.pos='H'}
bind_rows(
  genk %>% summarise(Group = "All cases",
                     Count = sum(planned_over, na.rm = TRUE),
                     Percentage = round(100 * mean(planned_over, na.rm = TRUE), 1)),
  genk %>% group_by(Group = urgency_lbl) %>%
    summarise(Count = sum(planned_over, na.rm = TRUE),
              Percentage = round(100 * mean(planned_over, na.rm = TRUE), 1))
) %>%
  pander(split.table = 120,
         caption = "Cases planned to end past the shift boundary. \\label{tab:pov}")
```

# How far past the boundary are these cases planned?

The planned excess is substantial rather than marginal. Among planned crossings, the median planned end lies 40 minutes past the boundary and the 95th percentile 246 minutes past it. The distribution spreads across all severities: roughly a fifth of planned crossings are planned 15 minutes or less past the boundary, while about a third are planned more than an hour past it. Table \ref{tab:bins} groups the cases by planned excess and Figure \ref{fig:povhist} shows the distribution.

```{r bins, table.pos='H'}
genk %>% filter(planned_over) %>%
  mutate(Bin = cut(pov_min, c(0, 15, 30, 60, 120, Inf),
                   labels = c("1-15 min", "16-30 min", "31-60 min", "61-120 min", ">120 min"))) %>%
  count(Bin, urgency_lbl) %>%
  pivot_wider(names_from = urgency_lbl, values_from = n, values_fill = 0) %>%
  mutate(Total = Elective + Urgent,
         Percentage = round(100 * Total / sum(Total), 1)) %>%
  pander(split.table = 120,
         caption = "Planned minutes past the shift boundary among planned crossings. \\label{tab:bins}")
```

```{r povhist, fig.cap="Planned minutes past the shift boundary (planned crossings, trimmed at P99) \\label{fig:povhist}", fig.pos='H'}
upper <- genk %>% filter(planned_over) %>% pull(pov_min) %>% quantile(.99, na.rm = TRUE)
genk %>% filter(planned_over, pov_min <= upper) %>%
  ggplot(aes(x = pov_min)) +
  geom_histogram(binwidth = 15, fill = "steelblue", color = "white") +
  labs(title = "Planned minutes past the shift boundary",
       x = "Planned minutes past boundary", y = "Number of surgeries") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

# Do planned crossings become actual overtime?

They largely do. Of the 5,812 planned crossings, 60.8% actually ran past the boundary, against 5.7% of the cases that were planned to fit within their shift. Overtime risk is therefore visible at planning time: a case planned to cross the boundary is more than ten times as likely to end in overtime as a case planned to fit. Table \ref{tab:funnel} summarises the conversion.

```{r funnel, table.pos='H'}
bind_rows(
  genk %>% filter(planned_over) %>%
    summarise(Group = "Planned to cross", Cases = n(),
              `Ran overtime` = sum(afterhours_flag == 1, na.rm = TRUE),
              `Overtime rate (%)` = round(100 * mean(afterhours_flag == 1, na.rm = TRUE), 1)),
  genk %>% filter(!planned_over) %>%
    summarise(Group = "Planned to fit", Cases = n(),
              `Ran overtime` = sum(afterhours_flag == 1, na.rm = TRUE),
              `Overtime rate (%)` = round(100 * mean(afterhours_flag == 1, na.rm = TRUE), 1))
) %>%
  pander(split.table = 120,
         caption = "Conversion of planned crossings into observed overtime. \\label{tab:funnel}")
```

# Decomposing observed overtime

The 7,729 overtime cases divide into two components. Scheduled crossings, planned past the boundary from the start, account for 3,532 cases (45.7%). Unplanned overruns, planned to fit but running long, account for 4,197 cases (54.3%). The scheduled component runs considerably heavier: a mean of 82 overtime minutes against 42 for unplanned overruns, so scheduled crossings carry 62.1% of all overtime minutes despite being the smaller group of cases. Table \ref{tab:decomp} reports the split.

```{r decomp, table.pos='H'}
ot %>%
  mutate(Kind = ifelse(planned_over, "Scheduled crossing", "Unplanned overrun")) %>%
  group_by(Kind) %>%
  summarise(Cases = n(),
            `Share of overtime cases (%)` = round(100 * n() / nrow(ot), 1),
            `Mean OT (min)` = round(mean(overtime_minutes), 1),
            `Median OT (min)` = round(median(overtime_minutes), 1),
            `Total OT minutes` = round(sum(overtime_minutes))) %>%
  mutate(`Share of OT minutes (%)` = round(100 * `Total OT minutes` / sum(`Total OT minutes`), 1)) %>%
  pander(split.table = 140,
         caption = "Scheduled versus unplanned overtime. \\label{tab:decomp}")
```

# Which kind of overtime does each room produce?

The balance between the two components differs strongly between rooms. In GO10, the room with the highest overtime rate, 81.5% of overtime cases were scheduled crossings: its overtime is predominantly written into the schedule, consistent with long cardiac procedures that cannot fit inside the day shift. In most other rooms the two components are closer to balanced. Table \ref{tab:rooms} decomposes each room's overtime rate and Figure \ref{fig:stacked} shows the same split graphically.

```{r rooms, table.pos='H'}
room_tbl <- genk %>% group_by(ActualOR) %>%
  summarise(Cases = n(),
            `OT rate (%)` = round(100 * mean(afterhours_flag == 1, na.rm = TRUE), 1),
            `Scheduled (%)` = round(100 * mean(planned_over & afterhours_flag == 1, na.rm = TRUE), 1),
            `Unplanned (%)` = round(100 * mean(!planned_over & afterhours_flag == 1, na.rm = TRUE), 1)) %>%
  arrange(desc(`OT rate (%)`))
room_tbl %>%
  pander(split.table = 120,
         caption = "Room-level overtime decomposed into scheduled and unplanned components. \\label{tab:rooms}")
```

```{r stacked, fig.cap="Room-level overtime rate split into scheduled and unplanned components \\label{fig:stacked}", fig.pos='H'}
room_tbl %>%
  select(ActualOR, Scheduled = `Scheduled (%)`, Unplanned = `Unplanned (%)`) %>%
  pivot_longer(-ActualOR, names_to = "Component", values_to = "Rate") %>%
  ggplot(aes(x = reorder(ActualOR, Rate, sum), y = Rate, fill = Component)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c(Scheduled = "steelblue", Unplanned = "grey70")) +
  labs(title = "Overtime rate per room, by component",
       x = "Operating room", y = "Overtime rate (%)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

# Concentration of overtime minutes

Overtime minutes are strongly concentrated. GO10 alone, with 2.2% of the caseload, generates 19.0% of all overtime minutes; together with GO11 and GO05 the three largest contributors account for 36.4%. Table \ref{tab:conc} lists the five rooms with the largest contributions.

```{r concentration, table.pos='H'}
tot_min <- sum(ot$overtime_minutes)
ot %>% group_by(ActualOR) %>%
  summarise(`OT minutes` = round(sum(overtime_minutes))) %>%
  mutate(`Share of all OT minutes (%)` = round(100 * `OT minutes` / tot_min, 1)) %>%
  arrange(desc(`OT minutes`)) %>%
  slice_head(n = 5) %>%
  pander(split.table = 120,
         caption = "Rooms contributing the most overtime minutes. \\label{tab:conc}")
```

# Room-level correlations

For completeness, the two room-level rank correlations are recomputed across the 18 rooms. Late-start rate shows no association with overtime rate (rho = -0.29, p = 0.24). Mean idle time between cases correlates strongly with overtime rate, whether the room rate is computed over all cases (rho = 0.91) or restricted to cases with a defined gap (rho = 0.89), and the association persists when GO10 is excluded (rho = 0.90). Table \ref{tab:corr} summarises the four analyses.

```{r correlations, table.pos='H'}
room <- genk %>% group_by(ActualOR) %>%
  summarise(late = mean(start_diff > 0, na.rm = TRUE),
            otr  = mean(afterhours_flag == 1, na.rm = TRUE),
            gap  = mean(gap_time, na.rm = TRUE))
sub <- genk %>% filter(!is.na(gap_time)) %>% group_by(ActualOR) %>%
  summarise(gap = mean(gap_time), otr = mean(afterhours_flag == 1))
no10 <- room %>% filter(ActualOR != "GO10")

ct <- function(x, y) cor.test(x, y, method = "spearman")
res <- list(
  `Late-start rate vs OT rate`        = ct(room$late, room$otr),
  `Mean idle vs OT rate (all cases)`  = ct(room$gap,  room$otr),
  `Mean idle vs OT rate (gap subset)` = ct(sub$gap,   sub$otr),
  `Mean idle vs OT rate (excl. GO10)` = ct(no10$gap,  no10$otr)
)
tibble(Analysis = names(res),
       rho = round(sapply(res, function(z) unname(z$estimate)), 3),
       `p value` = signif(sapply(res, function(z) z$p.value), 3)) %>%
  pander(split.table = 120,
         caption = "Spearman rank correlations across the 18 rooms. \\label{tab:corr}")
```

# Conclusion

The schedule itself already crosses the shift boundary for 7.3% of cases, and these scheduled crossings account for nearly half of all overtime cases and about two thirds of all overtime minutes. Overtime at Genk therefore has two components with different management levers: scheduled crossings, which reflect planning and capacity decisions and dominate in the highest-overtime room, and unplanned overruns, which reflect duration variability. Overtime risk is largely visible at planning time, since cases planned to cross the boundary ran overtime more than ten times as often as cases planned to fit. Whether planning past the boundary reflects deliberate use of the evening staffing capacity is a question to take to the clinical team.
